Towards a free/open-source universal-dependency treebank for Kazakh

### 2015-08-26 2015-turklang-kazdep/paper.tex
table of functions associated with each case
git-svn-id: https://svn.code.sf.net/p/apertium/svn/branches/papers/@61589 72bbbca6-d526-0410-a7d9-f06f51895060

%\makeatletter
%\let\mcnewpage=\newpage
%\newcommand{\TrickSupertabularIntoMulticols}{%
  %\renewcommand\newpage{%
    %\if@firstcolumn
      %\hrule width\linewidth height0pt
      %\columnbreak
    %\else
      %\mcnewpage
    %\fi
  %}%
%}
%\makeatother

@@

%%%%% Stuff for this paper %%%%%
\usepackage{tikz-dependency}
%\newcommand{\gmk}[1]{{\rm {\sc \texttt{#1}}}}
\newcommand{\gmk}[1]{{\rm {\ll \textsc{#1}}}}
\newcommand{\kazakh}[1]{{\em #1}}
\newcommand{\gloss}[1]{`#1'}
\newcommand{\tgloss}[1]{{\em #1}}
%\newcommand{\tag}[1]{\texttt{‹#1›}}
\newcommand{\tag}[1]{{\ll \textsc{‹#1›}}}
%\newcommand{\udtag}[1]{\texttt{@#1}}
\newcommand{\udtag}[1]{{\ll \textsc{#1}}}
\newcommand{\udlabel}[1]{{\udfont #1}}
\newcommand{\tilda}{{\fsfont ∼}}
%\newcommand{\corrauthor}[1]{\footnotemark\footnotetext{hargle}}

@@

\begin{table}[htbp]
	\centering
	\caption{Primary syntactic functions of traditionally defined cases in Kazakh.}\label{tab:cases}
	\begin{small}
		\begin{tabular}{l l p{9em} p{27.15em}}
			\toprule
				\textbf{Case} & \textbf{Morph} & \textbf{Functions} & \textbf{Examples} \\
			\midrule
				nom. & - & subject\newline
										attributive\newline
										indefinite object\newline
										indefinite genitival 
									& \kazakh{\textbf{Дәрігер} үйді көреді.} \gloss{\textbf{The doctor} sees the house.} \newline
										\kazakh{\textbf{қонақ} үй} \gloss{hotel (lit., \textbf{guest} house)} \newline 
										\kazakh{Дәрігер \textbf{үй} көреді.} \gloss{The doctor sees \textbf{a house}.}\newline
										\kazakh{\textbf{үй} жануарлары} \gloss{\textbf{house} animals} \\\midrule
				acc. & -/NI/ & definite object %\newline
										%	embedded subject\newline
										%	intr.\ causative subj.
										& \kazakh{Дәрігер \textbf{үйді} көреді.} \gloss{The doctor sees \textbf{the house}.}
										%	bar
										\\\midrule
				gen. & -/NIң/ & definite genitival\newline
											embedded subject
										& \kazakh{\textbf{үйдің} жануарлары} \gloss{the animals \textbf{of the house}} \newline
											\kazakh{Ол \textbf{Айгүлдің} ойнағанына қарап тұр.} \gloss{She's watching \textbf{Aygül} play.} \\\midrule
				loc. & -/DA/ & adverbial
										& \kazakh{\textbf{Үйде} ұйықтап жатыр.} \gloss{S/he's sleeping \textbf{in the house}.} \\\midrule
				abl. & -/DAн/ & adverbial\newline
											comparator 
										& \kazakh{Дәрігер \textbf{үйден} шықты.} \gloss{The doctor came out \textbf{of the house}.}\newline
											\kazakh{\textbf{түйеден} үлкен} \gloss{bigger \textbf{than a camel}}\\\midrule
				dat. & -/GA/ & indirect object\newline
										adverbial\newline
									%	attributive\newline
										trans.\ causative subj.%\newline
									%	nom.\ complement
									& \kazakh{\textbf{Ініме} кітап бердім.} \gloss{I gave \textbf{my younger brother} a book.} \newline
										\kazakh{\textbf{Тойға} көп кісі келіпті.} \gloss{A lot of people came \textbf{to the feast.}}\newline
										\kazakh{\textbf{Балаға} әнді тыңдаттым.} \gloss{I made \textbf{the child} listen to the song.}
										\\\midrule
				inst. & -/Mен/ &  adverbial & \kazakh{Олар \textbf{күшікпен} ойнап тұр.} \gloss{They're playing \textbf{with the dog}.} \\
			\bottomrule
		\end{tabular}
	\end{small}
\end{table}

@@

\begin{table}[htbp]
	\centering
	\caption{The default (first line of each) and ``zero-derived'' uses of nominals, adjectives, and adverbs}\label{tab:zeroderiv}
	\begin{tabular}{lp{5em}p{20em}}
		\toprule
			\textbf{Category} & \textbf{Function} & \textbf{Example} \\
		\midrule
			Nominals & Substantive\newline
							Attributive\newline
							Adverbial
						& \kazakh{\textbf{Қонақтарымыз} бай.} \gloss{\textbf{Our guests} are rich.}\newline 
							\kazakh{\textbf{қонақ} үй} \gloss{hotel (lit., \textbf{guest} house)}\newline
							\kazakh{Оны бір \textbf{рет} көрдім.} \gloss{I saw him/her/it one \textbf{time}.} \\\midrule
			Adjectives & Attributive\newline
							Substantive\newline
							Adverbial
						& \kazakh{\textbf{жақсы} үй} \gloss{\textbf{nice} house} \newline
						\kazakh{ең \textbf{жақсы}лары} \gloss{\textbf{the nice}st \textbf{one}s} \newline
						\kazakh{Оны \textbf{жақсы} танимын.} \gloss{I know him/her \textbf{well}.}
						\\\midrule
			Adverbs & Adverbial
						& \kazakh{Ол \textbf{кеше} кетті.} \gloss{S/he left \textbf{yesterday}.}
						\\
		\bottomrule
	\end{tabular}

@@

\begin{table}[htbp]
  \centering
  \caption{Composition of the corpus. The corpus covers a range of genres and text types from free \& public-domain sources.}
  \label{table:corpcomp}
	\begin{small}
		\begin{tabular}{llrrr}
			\toprule
				\textbf{Document} & \textbf{Description} & \textbf{Sentences} & \textbf{Tokens} & \textbf{Avg. length}\\
			\midrule
				UN Declaration on Human Rights & Legal text on human rights & 26 & 417 & 16.0 \\ %
				Phrasebook                     & Phrases from Wikitravel    & 38 & 204 & 5.4 \\ %
				Жиырма Бесінші Сөз             & Philosophical text         & 33 & 467 & 14.2 \\ %
				Қожанасырдың тойға баруы       & Folk tale from Wikisource  & 9 & 139 & 15.4 \\ %
				Ер Төстік                      & Folk tale from Wikisource  & 25 & 209 & 8.4 \\ %
				Азамат қайда?                  & Story for language learners & 49 & 433 & 8.8 \\ %
				Футболдан әлем чемпионаты 2014 & Wikipedia article (2014 World Cup) & 12 & 191 & 15.9 \\ %
				Иран & Wikipedia article (Iran)                             & 121 & 1714 & 14.2 \\ %
				Радиан & Wikipedia article (Radian)                         & 2 & 17 & 8.5 \\ %
				Шымкент & Wikipedia article (Shymkent)                      & 13 & 160 & 12.3 \\ %
				Wikipedia misc. & Random sentences from Wikipedia           & 74 & 565 & 7.6 \\ %
			\midrule
			\multicolumn{2}{c}{~} & 402 & 4516 & 11.2 \\\cmidrule[\heavyrulewidth]{3-5}
		\end{tabular}
	\end{small}
\end{table}
\vspace{-0.5em}

@@

        \begin{subfigure}[b]{0.25\textwidth}
                \centering
                \begin{dependency}[theme = simple, font = \small]
                   \begin{deptext}[column sep=0.2cm]
                                Айгүл \& студент \& · \\
                                \gmk{np} \& \gmk{n} \& \gmk{aor} \\
                                \tgloss{Aygül} \& \tgloss{student} \& \tgloss{is} \\
                        \end{deptext}
                        \deproot[edge unit distance=1.3ex]{2}{\udlabel{root}}
                        \depedge{2}{1}{\udlabel{subj}}
                        \depedge{2}{3}{\udlabel{cop}}
                \end{dependency}
                \caption{Non-surfaced copula.}\label{fig:cop1}
        \end{subfigure}
        \quad
        \begin{subfigure}[b]{0.25\textwidth}
                \centering
                \begin{dependency}[theme = simple, font = \small]
                   \begin{deptext}[column sep=0.2cm]
                                Айгүл \& студент \& болған \\
                                \gmk{np} \& \gmk{n} \& \gmk{past} \\
                                \tgloss{Aygül} \& \tgloss{student} \& \tgloss{was}\\
                        \end{deptext}
                        \deproot[edge unit distance=1.3ex]{2}{\udlabel{root}}
                        \depedge{2}{1}{\udlabel{subj}}
                        \depedge{2}{3}{\udlabel{cop}}
                \end{dependency}
                \caption{Copula in past tense.}\label{fig:cop2}
        \end{subfigure}
        \quad
        \begin{subfigure}[b]{0.40\textwidth}
                \centering
                \begin{dependency}[theme = simple, font = \small]
                   \begin{deptext}[column sep=0.2cm]
                                Айгүл \& оның \& қайда \& екенін \& білмейді. \\
                                \gmk{np} \& \gmk{gen} \& \gmk{adv} \& \gmk{cop} \& \gmk{neg.aor} \\
                                \tgloss{Aygül} \& \tgloss{his} \& \tgloss{where} \& \tgloss{being} \& \tgloss{knows not} \\
                        \end{deptext}
                        \deproot[edge unit distance=1.3ex]{5}{\udlabel{root}}
                        \depedge{3}{2}{\udlabel{subj}}
                        \depedge{5}{3}{\udlabel{ccomp}}
                        \depedge{3}{4}{\udlabel{cop}}
                \end{dependency}
                \caption{Copula in an embedded clause.}\label{fig:cop3}
        \end{subfigure}
        \caption{Dependency trees of copula constructions.}
\end{figure}
\vspace{-0.5em}

@@

As mentioned earlier, verbal nouns combined with certain case morphology and/or postpositions can function as an adverbial adjunct to a main clause, similar to verbal adverb clauses.  In such instances, they receive the UD label \udtag{advcl}, and example of which is given in figure~\ref{fig:gerabl}.

\begin{figure}[htbp]
	\centering
	\begin{dependency}[theme = simple, font = \small]
		\begin{deptext}[column sep=0.08cm]
			Айгүл \& санап \& біткеннен \& кейін \& айналасына \& қарады. \\
			\gmk{n.nom} \& \gmk{v.iv.prc} \& \gmk{vaux.ger.abl} \& \gmk{post} \& \gmk{n.px3sp.dat} \& \gmk{v.tv.ifi} \\
			\tgloss{Aygül} \& \tgloss{counting} \& \tgloss{having finished} \& \tgloss{after} \& \tgloss{to her surroundings} \& \tgloss{looked} \\
		\end{deptext}
		\depedge{6}{2}{\udlabel{advcl}}
		\depedge{2}{3}{\udlabel{aux}}
		\depedge{3}{4}{\udlabel{post}}
		\depedge{6}{5}{\udlabel{nmod}}
		\deproot[edge unit distance=1.3ex]{6}{\udlabel{root}}
	\end{dependency}
	\caption{Verbal nouns with case/postpositions: verbal nouns with certain case morphology and/or postpositions are the head of everything in their own clause and can be an \udtag{advcl} dependency of the verb they are the complement of.}\label{fig:gerabl}
\end{figure}

>>>>

%\makeatletter
%\let\mcnewpage=\newpage
%\newcommand{\TrickSupertabularIntoMulticols}{%
  %\renewcommand\newpage{%
    %\if@firstcolumn
      %\hrule width\linewidth height0pt
      %\columnbreak
    %\else
      %\mcnewpage
    %\fi
  %}%%
%}
%\makeatother

@@

%%%%% Stuff for this paper %%%%%
\usepackage{tikz-dependency}
%\newcommand{\gmk}[1]{{\rm {\sc \texttt{#1}}}}
\newcommand{\gmk}[1]{{\rm {\ll \textsc{#1}}}}
\newcommand{\kazakh}[1]{{\em #1}}
\newcommand{\gloss}[1]{`#1'}
\newcommand{\sgloss}[1]{\hspace{0.5em}`#1'}
%\newcommand{\sgloss}[1]{\hfill{}`#1'}
\newcommand{\tgloss}[1]{{\em #1}}
%\newcommand{\tag}[1]{\texttt{‹#1›}}
\newcommand{\tag}[1]{{\ll \textsc{‹#1›}}}
%\newcommand{\udtag}[1]{\texttt{@#1}}
\newcommand{\udtag}[1]{{\ll \textsc{#1}}}
\newcommand{\udlabel}[1]{{\udfont #1}}
\newcommand{\tilda}{{\fsfont ∼}}
%\newcommand{\corrauthor}[1]{\footnotemark\footnotetext{hargle}}

@@

\begin{table}[htbp]
	\centering
	\caption{Primary syntactic functions of traditionally defined cases in Kazakh.}\label{tab:cases}
	\begin{small}
		\begin{tabular}{l l p{9em} p{28em}}
			\toprule
				\textbf{Case} & \textbf{Morph} & \textbf{Functions} & \textbf{Examples} \\
			\midrule
				nom. & - & subject\newline
										attributive\newline
										indefinite object\newline
										indefinite genitival 
									& \kazakh{\textbf{Дәрігер} үйді көреді.} \sgloss{\textbf{The doctor} sees the house.} \newline
										\kazakh{\textbf{қонақ} үй} \sgloss{hotel (lit., \textbf{guest} house)} \newline 
										\kazakh{Дәрігер \textbf{үй} көреді.} \sgloss{The doctor sees \textbf{a house}.}\newline
										\kazakh{\textbf{үй} жануарлары} \sgloss{\textbf{house} animals} \\\midrule
				acc. & -/NI/ & definite object %\newline
										%	embedded subject\newline
										%	intr.\ causative subj.
										& \kazakh{Дәрігер \textbf{үйді} көреді.} \sgloss{The doctor sees \textbf{the house}.}
										%	bar
										\\\midrule
				gen. & -/NIң/ & definite genitival\newline
											embedded subject
										& \kazakh{\textbf{үйдің} жануарлары} \sgloss{the animals \textbf{of the house}} \newline
											\kazakh{Ол \textbf{Айгүлдің} ойнағанына қарап тұр.} \sgloss{She's watching \textbf{Aygül} play.} \\\midrule
				loc. & -/DA/ & adverbial
										& \kazakh{\textbf{Үйде} ұйықтап жатыр.} \sgloss{S/he's sleeping \textbf{in the house}.} \\\midrule
				abl. & -/DAн/ & adverbial\newline
											comparator 
										& \kazakh{Дәрігер \textbf{үйден} шықты.} \sgloss{The doctor came out \textbf{of the house}.}\newline
											\kazakh{\textbf{түйеден} үлкен} \sgloss{bigger \textbf{than a camel}}\\\midrule
				dat. & -/GA/ & indirect object\newline
										adverbial\newline
									%	attributive\newline
										trans.\ causative subj.%\newline
									%	nom.\ complement
									& \kazakh{\textbf{Ініме} кітап бердім.} \sgloss{I gave \textbf{my younger brother} a book.} \newline
										\kazakh{\textbf{Тойға} көп кісі келіпті.} \sgloss{A lot of people came \textbf{to the feast.}}\newline
										\kazakh{\textbf{Балаға} әнді тыңдаттым.} \sgloss{I made \textbf{the child} listen to the song.}
										\\\midrule
				inst. & -/Mен/ &  adverbial & \kazakh{Олар \textbf{күшікпен} ойнап тұр.} \sgloss{They're playing \textbf{with the dog}.} \\
			\bottomrule
		\end{tabular}
	\end{small}
\end{table}

@@

\begin{table}[htbp]
	\centering
	\caption{The default (first line of each) and ``zero-derived'' uses of nominals, adjectives, and adverbs}\label{tab:zeroderiv}
	{\small
	\begin{tabular}{lp{5em}p{21em}}
		\toprule
			\textbf{Category} & \textbf{Function} & \textbf{Example} \\
		\midrule
			Nominals & Substantive\newline
							Attributive\newline
							Adverbial
						& \kazakh{\textbf{Қонақтарымыз} бай.} \sgloss{\textbf{Our guests} are rich.}\newline 
							\kazakh{\textbf{қонақ} үй} \sgloss{hotel (lit., \textbf{guest} house)}\newline
							\kazakh{Оны бір \textbf{рет} көрдім.} \sgloss{I saw him/her/it one \textbf{time}.} \\\midrule
			Adjectives & Attributive\newline
							Substantive\newline
							Adverbial
						& \kazakh{\textbf{жақсы} үй} \sgloss{\textbf{nice} house} \newline
						\kazakh{ең \textbf{жақсы}лары} \sgloss{\textbf{the nice}st \textbf{one}s} \newline
						\kazakh{Оны \textbf{жақсы} танимын.} \sgloss{I know him/her \textbf{well}.}
						\\\midrule
			Adverbs & Adverbial
						& \kazakh{Ол \textbf{кеше} кетті.} \sgloss{S/he left \textbf{yesterday}.}
						\\
		\bottomrule
	\end{tabular}
	}

@@

\begin{table}[htbp]
  \centering
  \caption{Composition of the corpus. The corpus covers a range of genres and text types from free \& public-domain sources.}
  \label{table:corpcomp}
	\begin{small}
		\begin{tabular}{llrrr}
			\toprule
				\textbf{Document} & \textbf{Description} & \textbf{Sentences} & \textbf{Tokens} & \textbf{Avg. length}\\
			\midrule
				UN Declaration on Human Rights & Legal text on human rights & 26 & 417 & 16.0 \\ %
				Phrasebook                     & Phrases from Wikitravel    & 38 & 204 & 5.4 \\ %
				Жиырма Бесінші Сөз             & Philosophical text         & 33 & 467 & 14.2 \\ %
				Қожанасырдың тойға баруы       & Folk tale from Wikisource  & 9 & 139 & 15.4 \\ %
				Ер Төстік                      & Folk tale from Wikisource  & 25 & 209 & 8.4 \\ %
				Азамат қайда?                  & Story for language learners & 49 & 433 & 8.8 \\ %
				Футболдан әлем чемпионаты 2014 & Wikipedia article (2014 World Cup) & 12 & 191 & 15.9 \\ %
				Иран & Wikipedia article (Iran)                             & 121 & 1714 & 14.2 \\ %
				Радиан & Wikipedia article (Radian)                         & 2 & 17 & 8.5 \\ %
				Шымкент & Wikipedia article (Shymkent)                      & 13 & 160 & 12.3 \\ %
				Wikipedia misc. & Random sentences from Wikipedia           & 74 & 565 & 7.6 \\ %
			\midrule
			\multicolumn{2}{c}{~} & 402 & 4516 & 11.2 \\\cmidrule[\heavyrulewidth]{3-5}
		\end{tabular}
	\end{small}
\end{table}

Table~\ref{tab:casefunc} details the various tags (i.e., syntactic functions) associated with the traditionally defined morphological cases of Kazakh.  It should be noted that there are several known uses which are currently unattested (partitive use of \gmk{abl} as \udtag{dobj}, embedded \udtag{subj} marked as \gmk{acc}) or rare (\gmk{dat} as \udtag{iobj}) in the corpus.

\begin{table}[htbp]
	\centering
	\caption{The functions associated with each case in Kazakh as currently found in the corpus.}
	\label{tab:casefunc}
	\begin{tabular}{lrrrrrrrr}
		\toprule
			 & \udtag{iobj} & \udtag{cmpnd} & \udtag{nmod} & \udtag{nmod:poss} & \udtag{dobj} & \udtag{root} & \udtag{subj} & \udtag{voc} \\
		\midrule
			\gmk{nom} & — & 1 & 199 & 158 & 72 & 26 & 279 & 1 \\
			\gmk{acc} & — & — & 1 & — & 101 & — & — & — \\
			\gmk{gen} & — & — & 120 & 120 & — & — & 12 & — \\
			\gmk{dat} & 1 & — & 113 & — & 13 & — & — & — \\
			\gmk{loc} & — & — & 112 & 1 & — & 6 & — & — \\
			\gmk{abl} & — & — & 64 & — & — & — & — & — \\
			\gmk{ins} & — & — & 33 & — & — & — & — & — \\
		\bottomrule
	\end{tabular}
\end{table}
\vspace{-0.75em}

@@

        \begin{subfigure}[b]{0.25\textwidth}
                \centering
                \begin{dependency}[theme = simple, font = \small]
                   \begin{deptext}[column sep=0.2cm]
                                Айгүл \& студент \& · \\
                                \gmk{np} \& \gmk{n} \& \gmk{aor} \\
                                \tgloss{Aygül} \& \tgloss{student} \& \tgloss{is} \\
                        \end{deptext}
                        \deproot[edge unit distance=1.3ex]{2}{\udlabel{root}}
                        \depedge{2}{1}{\udlabel{subj}}
                        \depedge{2}{3}{\udlabel{cop}}
                \end{dependency}
                \caption{Non-surfaced copula.}\label{fig:cop1}
        \end{subfigure}
        \quad
        \begin{subfigure}[b]{0.25\textwidth}
                \centering
                \begin{dependency}[theme = simple, font = \small]
                   \begin{deptext}[column sep=0.2cm]
                                Айгүл \& студент \& болған \\
                                \gmk{np} \& \gmk{n} \& \gmk{past} \\
                                \tgloss{Aygül} \& \tgloss{student} \& \tgloss{was}\\
                        \end{deptext}
                        \deproot[edge unit distance=1.3ex]{2}{\udlabel{root}}
                        \depedge{2}{1}{\udlabel{subj}}
                        \depedge{2}{3}{\udlabel{cop}}
                \end{dependency}
                \caption{Copula in past tense.}\label{fig:cop2}
        \end{subfigure}
        \quad
        \begin{subfigure}[b]{0.40\textwidth}
                \centering
                \begin{dependency}[theme = simple, font = \small]
                   \begin{deptext}[column sep=0.2cm]
                                Айгүл \& оның \& қайда \& екенін \& білмейді. \\
                                \gmk{np} \& \gmk{gen} \& \gmk{adv} \& \gmk{cop} \& \gmk{neg.aor} \\
                                \tgloss{Aygül} \& \tgloss{his} \& \tgloss{where} \& \tgloss{being} \& \tgloss{knows not} \\
                        \end{deptext}
                        \deproot[edge unit distance=1.3ex]{5}{\udlabel{root}}
                        \depedge{3}{2}{\udlabel{subj}}
                        \depedge{5}{3}{\udlabel{ccomp}}
                        \depedge{3}{4}{\udlabel{cop}}
                \end{dependency}
                \caption{Copula in an embedded clause.}\label{fig:cop3}
        \end{subfigure}
        \caption{Dependency trees of copula constructions.}
\end{figure}
\vspace{-0.75em}

@@

As mentioned earlier, verbal nouns combined with certain case morphology and/or postpositions can function as an adverbial adjunct to a main clause, similar to verbal adverb clauses.  In such instances, they receive the UD label \udtag{advcl}, and example of which is given in figure~\ref{fig:gerabl}.
\vspace{-0.5em}

\begin{figure}[htbp]
	\centering
	\begin{dependency}[theme = simple, font = \small]
		\begin{deptext}[column sep=0.08cm]
			Айгүл \& санап \& біткеннен \& кейін \& айналасына \& қарады. \\
			\gmk{n.nom} \& \gmk{v.iv.prc} \& \gmk{vaux.ger.abl} \& \gmk{post} \& \gmk{n.px3sp.dat} \& \gmk{v.tv.ifi} \\
			\tgloss{Aygül} \& \tgloss{counting} \& \tgloss{having finished} \& \tgloss{after} \& \tgloss{to her surroundings} \& \tgloss{looked} \\
		\end{deptext}
		\depedge{6}{2}{\udlabel{advcl}}
		\depedge{2}{3}{\udlabel{aux}}
		\depedge{3}{4}{\udlabel{post}}
		\depedge{6}{5}{\udlabel{nmod}}
		\deproot[edge unit distance=1.3ex]{6}{\udlabel{root}}
	\end{dependency}
	\caption{Verbal nouns with case/postpositions: verbal nouns with certain case morphology and/or postpositions are the head of everything in their own clause and can be an \udtag{advcl} dependency of the verb they are the complement of.}\label{fig:gerabl}
\end{figure}
\vspace{-0.5em}
